# Readme.md
========================================================

## Assignment 1 - Getting and Cleaning Data

 Purpose: This readme file will explain how the original data was accessed
 and how the script works. This file should be read in conjunction with the
 accomanying CodeBook.md

 The source of data for this analysis is as follows:
 
  http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

 The source data included the following explanation 

"The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data."



 It is assumed the original zip file was unzipped as is into the subdirectory called 
 "UCI HAR Dataset" of the working directory.
  
 Read in the original test and training datasets.
 Next, combine their content with rbind into a new dataframe called 'x_cmpl'

```{r}

x_test = as.data.frame(read.table(file = "UCI HAR Dataset/test/X_test.txt", header = FALSE))
x_train = as.data.frame(read.table(file = "UCI HAR Dataset/train/X_train.txt", header = FALSE))
x_cmpl = rbind(x_test, x_train)

```

 Read in the original 'features" text file. Since it is read in as a column factor and I
 want to use this as the column headings for the test and training dataset created above, 
 one needs to transform this into a row vector. Only take the second row of the transformed
 vector (to get rid of the line numbers that appear in the first row).

 Use the entries as the column names in the combined test and training dataframe


```{r}
features = as.data.frame(read.table(file = "UCI HAR Dataset/features.txt", header = FALSE))
features = (t(features))[2,]
colnames(x_cmpl) <- features

```

 
 This prelimanry review is based on those measures of mean and std deviation collected in the
 dataset for the time domain signals (prefixed with a 't') only. That is, prior to any 
 Fast Fourier Transforms having been applied. Other columns may be reviewed in the future.

 In order to get eliminate the other columns, repeated use the 'grepl' function.
 The grepl function below is set up like this:  (A or B) and C 
 
 - A: keeps those column names with a string containing 'mean()'  (averages) 
 
 - B: keeps those column names with a string containing 'std()'   (std deviations) 
 
 - C: keeps those column names beginning with a lowercase 't'  (indicating time domain signals) 

 The logical vector 'mean_and_sd' contains the resulting TRUEs and FALSEs indicating those 
 rows that meet the (A or B) and C criteria

```{r}
mean_and_sd <- ((grepl('mean()', colnames(x_cmpl))) | ((grepl('std()', colnames(x_cmpl))))) &
          (grepl('^t', colnames(x_cmpl)))

```

Apply the mean_and_std logical vector against the columns of test and training dataframe (x_cmpl) only keeps the desired columns

```{r}
x_mean_std <- x_cmpl[ ,mean_and_sd]

```
 These lines use the global substitution command to 'gsub' to programatically cleanup the
 original column names (as read in through the 'features' file above).
 The headings are spread out by inserting periods, removing parentheses, etc.

```{r}
colnames(x_mean_std) <- gsub('Body', "Body.", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('Gravity', "Gravity.", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('-', "", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('mean\\()', ".mean", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('std\\()', ".stddev", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('X', ".X", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('Y', ".Y", colnames(x_mean_std))
colnames(x_mean_std) <- gsub('Z', ".Z", colnames(x_mean_std))

```

Read in the activity codes for the test and training datasets.Then rbind the two separate files. The column name for the single variable in the dataframe was also added.

```{r}
y_test = as.data.frame(read.table(file = "UCI HAR Dataset/test/y_test.txt", header = FALSE))
y_train = as.data.frame(read.table(file = "UCI HAR Dataset/train/y_train.txt", header = FALSE))
y_cmpl = rbind(y_test, y_train)
colnames(y_cmpl) <- "activ.number"
```

Read in the participant identifiers in the study and then rbind the test and training datasets. The column name for the single variable in the dataframe was also labeled.

```{r}
subj_test = as.data.frame(read.table(file = "UCI HAR Dataset/test/subject_test.txt", header = FALSE))
subj_train = as.data.frame(read.table(file = "UCI HAR Dataset/train/subject_train.txt", header = FALSE))
subj_cmpl = rbind(subj_test, subj_train)
colnames(subj_cmpl) <- "participant.number"
```

Read in the activity number and label identifiers.  Label column names.

```{r}
activ_labels = as.data.frame(read.table(file = "UCI HAR Dataset/activity_labels.txt", header = FALSE))
colnames(activ_labels) <- c("activ.number", "activity")
```

 Merge the test and training activity code dataframe with the activity labels dataframe 
 in order to get the actual labels such as "walking', 'laying', etc. into the mix

```{r}
y_cmpl <- (merge(activ_labels, y_cmpl, by = 'activ.number'))

```
 Create the final dataframe with activity labels, participant identifiers, readable column names, and the actual data for those chosen fields related to mean and standard deviation.

```{r}

all_cmpl <- cbind(y_cmpl[2], subj_cmpl, x_mean_std)
```

 In order to calculate the mean in each column by the separate activities, first split the 
 large dataframe ('all_cmpl') into six pieces--corresponding to the six activities

 This was done by calculating a logical variable, for example, 
 
     '[all_cmpl$activity == "WALKING",]'
 
 that tested whether the row in the large dataframe contained the activity label 'WALKING'.
 This results in TRUEs wherever the row had the 'WALKING' label and FALSE otherwise
 Applying this logical vector against the rows of the large dataframe, all_cmpl, serves
 to subset the 'all_cmpl' dataframe.

```{r}

all_cmpl_1 <- all_cmpl[all_cmpl$activity == "WALKING",]
all_cmpl_2 <- all_cmpl[all_cmpl$activity == "WALKING_UPSTAIRS",]
all_cmpl_3 <- all_cmpl[all_cmpl$activity == "WALKING_DOWNSTAIRS",]
all_cmpl_4 <- all_cmpl[all_cmpl$activity == "SITTING",]
all_cmpl_5 <- all_cmpl[all_cmpl$activity == "STANDING",]
all_cmpl_6 <- all_cmpl[all_cmpl$activity == "LAYING",]
```

 After the subsets are complete, calculate the means of the entries in each column only for
 those columns with numerical entries (3 through 42)

```{r}

all_cmpl_1_mean <- colMeans(all_cmpl_1[3:42])
all_cmpl_2_mean <- colMeans(all_cmpl_2[3:42])
all_cmpl_3_mean <- colMeans(all_cmpl_3[3:42])
all_cmpl_4_mean <- colMeans(all_cmpl_4[3:42])
all_cmpl_5_mean <- colMeans(all_cmpl_5[3:42])
all_cmpl_6_mean <- colMeans(all_cmpl_6[3:42])
```
 Assemble the final single 'tidy' dataset by row binding the various means calculated on 
 the above lines of codes. The last step is the specify the row names in 'tidy'.

```{r}
tidy <- rbind (all_cmpl_1_mean, all_cmpl_2_mean, all_cmpl_3_mean,
              all_cmpl_4_mean, all_cmpl_5_mean, all_cmpl_6_mean)
rownames(tidy) = c('WALKING','WALKING_UPSTAIRS','WALKING_DOWNSTAIRS','SITTING','STANDING','LAYING')

```
